---
authors:
- Hugo Authors
date: "`r Sys.Date()`"
excerpt: A plot showing the weekly approval rate for president Biden
hero: 
title: "Biden's Approval Margins"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy = FALSE,
  size = "small")

options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width = 6.75, 
  fig.height = 6.75,
  fig.align = "center")
```

```{r load libraries, include = FALSE}
library(tidyverse)
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
```

# Biden's Approval Margins

In this post, we will use data from <www.fivethirtyeight.com> on all polls that track the president Biden's approval rate in the US.
(https://projects.fivethirtyeight.com/biden-approval-ratings)

## Import data
```{r fix date and calculate net approval rate, cache=TRUE}
# Import approval polls data directly off fivethirtyeight website
approval_polllist <- read_csv('https://projects.fivethirtyeight.com/biden-approval-data/approval_polllist.csv') 

glimpse(approval_polllist)

# Use `lubridate` to fix dates, as they are given as characters.
# Use enddate as the date of the poll result
approval_polllist <- approval_polllist %>% 
  mutate(enddate = mdy(enddate))

```

## Create a plot
Using the data, we wish to calculate the average net approval rate (approve - disapprove) for each week since Biden got into office. \n
Then we would like to plot the net approval rate, along with its 95% confidence interval.

### Tidy data and calculate Confidence Interval
```{r net approval margin for Biden}
#tidy data and calculate CI using formula. 
net_approval <- approval_polllist %>% 
  #We concentrate only on subgroup "voters"
  filter(!is.na(subgroup), subgroup == "Voters") %>%
  #Use lubridate to get week number
  mutate(week = isoweek(enddate),
         net_approval_day = approve - disapprove) %>% 
  group_by(week) %>%
  summarise(mean_net_approval = mean(net_approval_day),
            sd_net_approval = sd(net_approval_day),
            count = n(),
            se_twitter = sd_net_approval / sqrt(count),
            t_critical = qt(0.975, count - 1),
            lower_ci =  mean_net_approval - t_critical*se_twitter,
            upper_ci = mean_net_approval + t_critical*se_twitter)

#Report the weekly net approval rate for Biden
net_approval %>% 
  knitr::kable(bootstrap_options = c ("striped","hover","condensed","responsive")) %>%
  kableExtra::kable_styling()
```

### Plot the net approval rate
```{r Biden net approval rate plot, fig.align="center", fig.height=18, fig.width=26}
ggplot(net_approval, 
       aes(x= week, 
           y= mean_net_approval)) +
  geom_line(color = "red")+
  geom_point(color = "red", size = 1)+
  geom_smooth(color = "blue",
              level = 0,
              size = 1)+
  #add orange line at zero
  geom_hline(yintercept=0, 
             color = "orange", 
             size = 2)+
  theme_minimal()+
  #add confidence band using calculated CI
  geom_ribbon(aes(ymin = lower_ci, 
                  ymax = upper_ci),
              alpha=0.3,
              fill = "grey",
              color = "red") + 
  labs(
    title = "Estimating Approval Margin (approve-disapprove) for Joe Biden",
    subtitle = "Weekly average approval of subgroup \"Voters\"",
    x = "Week of the year",
    y = "Average Approval Margin (approve-disapprove)")+
  scale_y_continuous(breaks=seq(-15,10,2.5), limits=c(-15,40))+
  scale_x_continuous(breaks=seq(0,40,13))+
  theme(
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size=22, face="bold"),
    axis.title.y = element_text(size=22, face="bold"),
    plot.title = element_text(size = 22, face="bold"),
    plot.subtitle = element_text(size=20, face="bold")
  )+
   annotate("text", x=19.5, y=20, label="2021", color = "#333333", size=8)

```

## Compare Confidence Intervals

We then compare the confidence intervals for `week 5` and `week 25` to see if there are any changes in Biden's net approval rate.
```{r compare week 5 and 25 confidence intervals for approval rate}
net_approval_5_25 <- net_approval %>% 
  filter(week %in% c(5, 25)) %>% 
  mutate(
    ci_width = upper_ci - lower_ci) %>% 
  select(week, lower_ci, upper_ci, ci_width)

net_approval_5_25 %>% 
  knitr::kable(bootstrap_options = c ("striped","hover","condensed","responsive")) %>%
  kableExtra::kable_styling()
  
```

## Analysis of the difference

From the results, we can clearly see that the confidence interval for Biden's net approval rate has been narrower from week 5 to week 25. \n

The standard deviation in approval ratings is much larger in week 5 than in week 25, which generates a higher standard error and consequently a wider confidence interval. We assume this is because as after Biden has been elected for a longer period of time in week 25 (almost half a year), voters would become more clear about their approval or disapproval to the president. After Americans took over 25-week time to evaluate their newly elected president, they would probably have a clearer attitude towards Biden's policy changes, administration and national strategies. These clearer perceptions then result in this decreasing variance in approval ratings, consequently a lower standard deviation and ultimately more narrow confidence intervals.